##Run Analysis R, a Script to Clean and Tidy Activity Data

The script performs the following tasks: 

1. merges the training and test sets to create one data set.
2. extracts the measurements on the mean and std dev for each measurement.
3. uses descriptive activity names to name the activities in the data set
4. appropriately labels the data set with descriptive variable names
5. from the data set in step 4, creates a second, independent tidy data set with the average of each variable for each activity and each subject.

Information about the dataset can be found at the repository:

http://archive.ics.uci.edu/ml/datasets/Human+Activity+Recognition+Using+Smartphones

###Merging the Data Sets

Merge the training and test sets to create one data set and #3, use descriptive activity names to name the activities in the data set, and #4 appropriately labels the data set with descriptive variable names.  The following script below accomplishes this task.

```{r}
X_train <- read.table("~/Downloads/UCI HAR Dataset/train/X_train.txt", quote="\"", stringsAsFactors=FALSE)

X_test <- read.table("~/Downloads/UCI HAR Dataset/test/X_test.txt", quote="\"", stringsAsFactors=FALSE)

library(tidyr)
library(dplyr)
library(reshape2)
library(data.table)

X_all <- bind_rows(X_train, X_test) #add all the training and test data together
View(X_all)
```

Now I name the columns something reasonable that indicates the data type.

```{r}
features <- read.table("~/Downloads/UCI HAR Dataset/features.txt", quote="\"", stringsAsFactors=FALSE)
View(features)
names(X_all) = features$V2

subject_train <- read.table("~/Downloads/UCI HAR Dataset/train/subject_train.txt", quote="\"", stringsAsFactors=FALSE)
subject_test <- read.table("~/Downloads/UCI HAR Dataset/test/subject_test.txt", quote="\"")

subject_all <- bind_rows(subject_train, subject_test) #combine as before in the x train and test sets

activity_labels <- read.table("~/Downloads/UCI HAR Dataset/activity_labels.txt", quote="\"", stringsAsFactors=FALSE)

View(activity_labels)

subject_activity <- full_join(subject_all, activity_labels) #descriptive activity names for all the activities in the data set on each subject

subject.no <- subject_activity$V1
activity <- subject_activity$V2
subject.activity <- data.frame(subject.no, activity)
all_data <- cbind(as.data.table(X_all),(subject_activity))
```

There are now two complete, corresponding data frames, one with all the X data and one with all the subject activity data by number and label

The next task is accomplished by joining the data to make one large data set.

```{r}
y_train <- read.table("~/Downloads/UCI HAR Dataset/train/y_train.txt", quote="\"", stringsAsFactors=FALSE)
y_test <- read.table("~/Downloads/UCI HAR Dataset/train/y_train.txt", quote="\"", stringsAsFactors=FALSE)

y_all <- bind_rows(y_train, y_test) #add all the training and test data together
View(y_all)
```

As a precautionary step, just for good measure, I create .csv output files to save and check data or to send to those who prefer other data analysis methods

```{r}
write.csv(all_data, "all_data.csv")
write.csv(subject_activity, "subject_activity.csv")
write.csv(y_all, "all_y_data.csv")
```

The next few scripts extracts measurements on the mean and std for each measurement; labels correctly with descriptive name. The first 6 columns in the all_data set have the mean and std, the rest are other measurements or quantiles.  Below is a new data set giving just these measurements.


```{r}
mean <- select(all_data, contains("mean"))
std <- select(all_data, contains("std"))
mean.std <- cbind(as.data.table(mean),(std))
mean.std.subject.activity <- cbind(as.data.table(mean.std), (subject.activity))

write.csv(mean.std.subject.activity, "mean.std.subject.activity.csv")
```

Create a tidy data set with average of each variable for each activity by each subject and save it as a text file.  Now you have a data file ready for further analysis.

```{r}
tidy <- na.omit(mean.std.subject.activity) #data set has no missing variables and contains all the means (averages) and std for each subject by activity in which they participated

write.table(tidy, row.name=FALSE, "tidy.txt")
```
##Code Book

Some General Information: Data come from "Human Activity Recognition database built from the recordings of 30 subjects performing activities of daily living (ADL) while carrying a waist-mounted smartphone with embedded inertial sensors." Which was supplied by UCI Center for Machine Learning and Intelligent Systems.

The master data set contains a multivariate, time-series type data of 10299 instances, for 561 attributes over 30 subjects. The source of the data is from smart-phone activity and provided by the individuals below:

Jorge L. Reyes-Ortiz(1,2), Davide Anguita(1), Alessandro Ghio(1), Luca Oneto(1) and Xavier Parra(2)
1 - Smartlab - Non-Linear Complex Systems Laboratory
DITEN - Università degli Studi di Genova, Genoa (I-16145), Italy. 
2 - CETpD - Technical Research Centre for Dependency Care and Autonomous Living
Universitat Politècnica de Catalunya (BarcelonaTech). Vilanova i la Geltrú (08800), Spain
activityrecognition '@' smartlab.ws

From the UCI site we have the following information:

Data Set Information:

The experiments have been carried out with a group of 30 volunteers within an age bracket of 19-48 years. Each person performed six activities (WALKING, WALKING_UPSTAIRS, WALKING_DOWNSTAIRS, SITTING, STANDING, LAYING) wearing a smartphone (Samsung Galaxy S II) on the waist. Using its embedded accelerometer and gyroscope, we captured 3-axial linear acceleration and 3-axial angular velocity at a constant rate of 50Hz. The experiments have been video-recorded to label the data manually. The obtained dataset has been randomly partitioned into two sets, where 70% of the volunteers was selected for generating the training data and 30% the test data. 

The sensor signals (accelerometer and gyroscope) were pre-processed by applying noise filters and then sampled in fixed-width sliding windows of 2.56 sec and 50% overlap (128 readings/window). The sensor acceleration signal, which has gravitational and body motion components, was separated using a Butterworth low-pass filter into body acceleration and gravity. The gravitational force is assumed to have only low frequency components, therefore a filter with 0.3 Hz cutoff frequency was used. From each window, a vector of features was obtained by calculating variables from the time and frequency domain.

A video of the experiment including an example of the 6 recorded activities with one of the participants can be seen in the following link: http://www.youtube.com/watch?v=XOEN9W05_4A.

###Attribute Information:

For each record in the dataset it is provided: 
- Triaxial acceleration from the accelerometer (total acceleration) and the estimated body acceleration. 
- Triaxial Angular velocity from the gyroscope. 
- A 561-feature vector with time and frequency domain variables. 
- Its activity label. 
- An identifier of the subject who carried out the experiment.

###Relevant Papers:

Davide Anguita, Alessandro Ghio, Luca Oneto, Xavier Parra and Jorge L. Reyes-Ortiz. Human Activity Recognition on Smartphones using a Multiclass Hardware-Friendly Support Vector Machine. International Workshop of Ambient Assisted Living (IWAAL 2012). Vitoria-Gasteiz, Spain. Dec 2012 

Davide Anguita, Alessandro Ghio, Luca Oneto, Xavier Parra, Jorge L. Reyes-Ortiz. Energy Efficient Smartphone-Based Activity Recognition using Fixed-Point Arithmetic. Journal of Universal Computer Science. Special Issue in Ambient Assisted Living: Home Care. Volume 19, Issue 9. May 2013

Davide Anguita, Alessandro Ghio, Luca Oneto, Xavier Parra and Jorge L. Reyes-Ortiz. Human Activity Recognition on Smartphones using a Multiclass Hardware-Friendly Support Vector Machine. 4th International Workshop of Ambient Assited Living, IWAAL 2012, Vitoria-Gasteiz, Spain, December 3-5, 2012. Proceedings. Lecture Notes in Computer Science 2012, pp 216-223. 

Jorge Luis Reyes-Ortiz, Alessandro Ghio, Xavier Parra-Llanas, Davide Anguita, Joan Cabestany, Andreu Català. Human Activity and Motion Disorder Recognition: Towards Smarter Interactive Cognitive Environments. 21th European Symposium on Artificial Neural Networks, Computational Intelligence and Machine Learning, ESANN 2013. Bruges, Belgium 24-26 April 2013.

###Citation Request:

Davide Anguita, Alessandro Ghio, Luca Oneto, Xavier Parra and Jorge L. Reyes-Ortiz. A Public Domain Dataset for Human Activity Recognition Using Smartphones. 21th European Symposium on Artificial Neural Networks, Computational Intelligence and Machine Learning, ESANN 2013. Bruges, Belgium 24-26 April 2013.

####Tidy Data Set Information

The tidy dataset, output called tidy.txt was created in the following manner:

One R script called run_analysis.R that does the following. 

1. Merges the training and the test sets to create one data set.
2. Extracts only the measurements on the mean and standard deviation for each measurement. 
3. Uses descriptive activity names to name the activities in the data set
4. Appropriately labels the data set with descriptive variable names. 
5. From the data set in step 4, creates a second, independent tidy data set with the average of each variable for each activity and each subject.

The descriptive variable names were taken from the original data set and modified so that they would provide information about the observation and variable.
